---
title: "Peer Assignmnet - Activity Monitoring Device"
author: "Niall Graham"
date: "16 December 2017"
output:
  html_document: default
  pdf_document: default
---
#Overview 
It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the “quantified self” movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks.

But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

#Data
The data for this assignment can be downloaded from the course web site.

The variables included in this dataset are:

  Steps: Number of steps taking in a 5-minute interval (missing values are coded as NA).
  Date: The date on which the measurement was taken in YYYY-MM-DD format.
  Interval: Identifier for the 5-minute interval in which measurement was taken.
  
The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

##Loading and preprocessing the data

Loading the Data
```{r}
echo=TRUE 
setwd("C:/Users/Niall Graham/Desktop/Data Science/Reproducible Data/Peer Grade assignment V2")
activity <- NULL
activity <- read.csv("activity.csv", header = T , sep = ",")
```

```{r}
echo = TRUE
df_summary <- NULL
su2 <- NULL
su <- NULL
mn_int <- NULL
activity2 <- NULL
mean_su2 <- NULL
median_su2 <- NULL
activity2_weekend <- NULL
activity2_weekday <- NULL
mean_activity2_weekday <- NULL
mean_activity2_weekend <- NULL
```

### Mean of  Total Number of Steps taken per day.

```{r}
echo = TRUE
su <- tapply(activity$steps, activity$date, sum, na.rm=T)
```

Histogram of the total number of steps taken each day
```{r}
echo = TRUE
hist(su, xlab = "Sum of Steps Per Day", main = "Histogram of Steps Per Day", col = "red")
```

Mean and the median total number of steps taken per day are reported :
```{r}
echo = TRUE
mean_su <- round(mean(su))
median_su <- round(median(su))
print(c("The mean is",mean_su))

print(c("The median is",median_su))
```


### The Average Daily Activity Pattern

Plot of the 5-minute interval and the average number of steps taken.

```{r}
echo = TRUE
mn_int <- tapply(activity$steps, activity$interval, mean, na.rm=T)
plot(mn_int ~ unique(activity$interval), type="l", xlab = "5-min Interval", col = "blue")
```

The 5-minute interval (on average across all the days in the dataset) that contains the maximum number of steps is the following: 

```{r}
echo = TRUE
mn_int[which.max(mn_int)]
```

##Imputing missing values

There are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

First, in order to visualize in which variable the NAs are:

```{r}
echo = TRUE 
table(is.na(activity) == TRUE)
```

```{r}
summary(activity)
```

There are 2304 NA's.

### Strategy for filling in all of the missing values in the dataset

```{r}
echo = TRUE
activity2 <- activity  # creation of the dataset that will have no more NAs
for (i in 1:nrow(activity)){
    if(is.na(activity$steps[i])){
        activity2$steps[i]<- mn_int[[as.character(activity[i, "interval"])]]
    }
}
```

Histogram of the Total Number of Steps Taken Each Day.

```{r}
echo = TRUE
su2 <- tapply(activity2$steps, activity2$date, sum, na.rm=T)
hist(su2, xlab = "sum of steps per day", main = "histogram of steps per day", col = "green")
```

```{r}
mean_su2 <- round(mean(su2))
median_su2 <- round(median(su2))
```

The new values are :
```{r}
echo = TRUE
print(c("The mean is",mean_su2))
```

```{r}
print(c("The median is",median_su2))
```

In order to compare the new values with the “old” values:
```{r}
echo = TRUE
df_summary <- rbind(df_summary, data.frame(mean = c(mean_su, mean_su2), median = c(median_su, median_su2)))
rownames(df_summary) <- c("with NA's", "without NA's")
print(df_summary)
```

For comparison with NA's and without (see earlier):
```{r}
echo = TRUE
summary(activity2)
```

###Are there differences in activity patterns between weekdays and weekends.

A new column is added to the dataframe, this column will contain the factor “weekday days”“ or "weekend days”.

```{r}
echo = TRUE
activity2$weekday <- c("weekday")
activity2[weekdays(as.Date(activity2[, 2])) %in% c("Saturday", "Sunday", "samedi", "dimanche", "saturday", "sunday", "Samedi", "Dimanche"), ][4] <- c("weekend")
table(activity2$weekday == "weekend")
```

```{r}
activity2$weekday <- factor(activity2$weekday)
```

In order to visualize the difference bewteen weekends and days of the week, a new dataframe is created to be usable by the lattice package. First, the data are calculated:

```{r}
echo = TRUE
activity2_weekend <- subset(activity2, activity2$weekday == "weekend")
activity2_weekday <- subset(activity2, activity2$weekday == "weekday")

mean_activity2_weekday <- tapply(activity2_weekday$steps, activity2_weekday$interval, mean)
mean_activity2_weekend <- tapply(activity2_weekend$steps, activity2_weekend$interval, mean)
```

```{r}
echo = TRUE
library(lattice)
df_weekday <- NULL
df_weekend <- NULL
df_final <- NULL
df_weekday <- data.frame(interval = unique(activity2_weekday$interval), avg = as.numeric(mean_activity2_weekday), day = rep("weekday", length(mean_activity2_weekday)))
df_weekend <- data.frame(interval = unique(activity2_weekend$interval), avg = as.numeric(mean_activity2_weekend), day = rep("weekend", length(mean_activity2_weekend)))
df_final <- rbind(df_weekday, df_weekend)

xyplot(avg ~ interval | day, data = df_final, layout = c(1, 2), 
       type = "l", ylab = "Number of steps")
```


